A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-03-01 sections/Metatheory.tex
S4 and Appendix lemma sync (minor swap of letters)
S4 double-checked Abella script against corollaries

\paragraph{Declarative transfer.}
Figure~\ref{fig:decl:worklist} defines a relation $\Om \rto \Om'$,
which transfers all judgements in the declarative worklists to the
declarative type system. This relation checks that every judgement entry in the worklist
holds using a corresponding conventional declarative judgement. 
The typing contexts of declarative judgements are recovered using an
auxiliary erasure function $\|\Om\|$. The erasure function
simply drops all judgment entries from the worklist, keeping only variable
and type variable declarations.

@@

\begin{lemma}[Subsumption]\label{lem:subsumption}
Given $\Psi' \le \Psi$:
\begin{enumerate}
    \item If $\Psi \vdash e \Lto A$ and $\Psi \vdash A \le A'$ then $\Psi' \vdash e \Lto A'$;
    \item If $\Psi \vdash e \To A$ then there exists
        $A'$ s.t. $\Psi \vdash A' \le A$ and $\Psi' \vdash e \To A'$;
    \item If $\Psi \vdash \appInf{C}{e}{A}$ and $\Psi \vdash C' \le C$,
        then there exists $A'$ s.t. $\Psi \vdash A' \le A$ and $\Psi' \vdash \appInf{C'}{e}{A'}$.
\end{enumerate}
\end{lemma}
This lemma expresses that any derivation in a context $\Psi$ has a corresponding derivation in any context
$\Psi'$ that it subsumes.

@@

\begin{corollary}[Soundness, single judgment form]
Given \emph{wf }$\Gm$:
\begin{enumerate}
    \item If $\Gm \Vdash A \le B \redto \nil$\\
        then there exist $A', B', \Om$ s.t.
        $\Gm \Vdash A \le B \sto \Om \Vdash A' \le B'$ and $\|\Om\| \vdash A' \le B'$;
    \item If $\Gm \Vdash e \Lto A \redto \nil$\\
        then there exist $ A', \Om$ s.t.
        $\Gm \Vdash e \Lto A \sto \Om \Vdash e \Lto A'$ and $\|\Om\| \vdash e \Lto A'$;
    \item If $\Gm \Vdash e \To_a \jg \redto \nil$ for any $\jg$\\
        then there exists $\Om, \jg', A$ s.t.
        $\Gm \sto \Om$ and $\|\Om\| \vdash e \To A$;
    \item If $\Gm \Vdash \appInfAlg{A}{e} \redto \nil$ for any $\jg$\\
        then there exists $\Om, \jg', A', C$ s.t.
        $\Gm \Vdash \appInfAlg{A}{e} \sto \Om \Vdash \appInfAlg{A'}{e}[a][\jg']$
            and $\|\Om\| \vdash \appInf{A'}{e}{C}$.
\end{enumerate}
\end{corollary}
\bruno{if possible double-check against the corresponding Abella ones.}

@@

The following corollary is derived immediately from Theorem~\ref{thm:completeness}.
\begin{corollary}[Completeness, single judgment form]
Given \emph{wf }$\Gm$ containing no judgment chains:
\begin{enumerate}
    \item If $\Om \vdash A' \le B'$ and $\Gm \Vdash A \le B \sto \Om \Vdash A' \le B'$
        then $\Gm \Vdash A \le B \redto \nil$;
    \item If $\Om \vdash e \Lto A'$ and $\Gm \Vdash e \Lto A \sto \Om \Vdash e \Lto A'$
        then $\Gm \Vdash e \Lto A \redto \nil$;
    \item If $\Om \vdash e \To A$ and $\Gm \sto \Om$
        then $\Gm \Vdash e \To_a 1 \le 1 \redto \nil$;
    \item If $\Om \vdash \appInf{A'}{e}{C}$ and
        $\Gm \Vdash \appInfAlg{A}{e}[a][1 \to 1] \sto \Om \Vdash \appInfAlg{A'}{e}[a][1 \to 1]$
        \\
        then $\Gm \Vdash \appInfAlg{A}{e}[a][1 \le 1] \redto \nil$.
\end{enumerate}
\end{corollary}
\bruno{We discussed the form of the lemmas on slack. Please have the first premise
  be a declarative judgement.
  If possible also double-check against Abella.}

@@

\begin{figure}[ht]
\centering\begin{tabular}{|c|c|c|l|}\hline
    File(s) & SLOC & \#Theorems & Description\\\hline
    olist.thm, nat.thm  &  309 & 56  & Basic data structures\\\hline
    typing.thm          &  238 & 5   & Declarative \& algorithmic system, debug examples\\\hline
    decl.thm            &  226 & 33  & Basic declarative properties\\\hline
    order.thm           &  235 & 27  & The $|\cdot|_\forall$ measure; decl. subtyping strenthening\\\hline
    alg.thm             &  655 & 72  & Basic algorithmic properties\\\hline
    trans.thm           &  628 & 53  & \makecell[l]{Transfer and its basic properties \\
                            Lemmas~\ref{lem:insert}, \ref{lem:extract}}\\\hline
    declTyping.thm      &  944 & 72  & \makecell[l]{Non-overlapping declarative system \\
                            Lemmas~\ref{lem:inv_allR}, \ref{lem:inv_chkAll}, \ref{lem:inv_chkLam}}\\\hline
    soundness.thm       & 1069 & 70  & Soundness theorem; aux. lemmas on transfer\\\hline
    depth.thm           &  206 & 14  & The $|\cdot|_\to$ measure; Lemma~\ref{lem:prune_inst}\\\hline
    dcl.thm             &  383 & 12  & Non-overlapping declarative worklist \\\hline
    completeness.thm    & 1075 & 61  &
                            Completeness theorem; aux. lemmas and relations\\\hline
    inst\_decidable.thm &  830 & 42  & Other measures; Lemma~\ref{lemma:inst:decidable}\\\hline
    decidability.thm    &  885 & 50  & Decidability theorem\\\hline
    Total               & 7683 & 567 & (51 definitions in total)\\\hline
\end{tabular}
\caption{Statistics for the proof scripts}
\label{fig:proof_statistics}
\end{figure}

>>>>

\paragraph{Declarative transfer.}
Figure~\ref{fig:decl:worklist} defines a relation $\Om \rto \Om'$,
which transfers all judgements in the declarative worklists to the
declarative type system. This relation checks that every judgement entry in the worklist
holds using a corresponding conventional declarative judgement. 
The typing contexts of declarative judgements are recovered using an
auxiliary erasure function $\|\Om\|$
\footnote{In the proof script we do not use the erasure function,
for the declarative system and well-formedness judgments automatically fits
the non-erased declarative worklist just as declarative contexts.}.
The erasure function simply drops all judgment entries from the worklist,
keeping only variable and type variable declarations.

@@

\begin{lemma}[Subsumption]\label{lem:subsumption}
Given $\Psi' \le \Psi$:
\begin{enumerate}
    \item If $\Psi \vdash e \Lto A$ and $\Psi \vdash A \le A'$ then $\Psi' \vdash e \Lto A'$;
    \item If $\Psi \vdash e \To B$ then there exists
        $B'$ s.t. $\Psi \vdash B' \le B$ and $\Psi' \vdash e \To B'$;
    \item If $\Psi \vdash \appInf{A}{e}{C}$ and $\Psi \vdash A' \le A$,
        then there exists $C'$ s.t. $\Psi \vdash C' \le C$ and $\Psi' \vdash \appInf{A'}{e}{C'}$.
\end{enumerate}
\end{lemma}
This lemma expresses that any derivation in a context $\Psi$ has a corresponding derivation in any context
$\Psi'$ that it subsumes.

@@

\begin{corollary}[Soundness, single judgment form]
Given \emph{wf }$\Gm$:
\begin{enumerate}
    \item If $\Gm \Vdash A \le B \redto \nil$\\
        then there exist $A', B', \Om$ s.t.
        $\Gm \Vdash A \le B \sto \Om \Vdash A' \le B'$ and $\|\Om\| \vdash A' \le B'$;
    \item If $\Gm \Vdash e \Lto A \redto \nil$\\
        then there exist $ A', \Om$ s.t.
        $\Gm \Vdash e \Lto A \sto \Om \Vdash e \Lto A'$ and $\|\Om\| \vdash e \Lto A'$;
    \item If $\Gm \Vdash e \To_a \jg \redto \nil$ for any $\jg$\\
        then there exists $\Om, \jg', A$ s.t.
        $\Gm \sto \Om$ and $\|\Om\| \vdash e \To A$;
    \item If $\Gm \Vdash \appInfAlg{A}{e} \redto \nil$ for any $\jg$\\
        then there exists $\Om, \jg', A', C$ s.t.
        $\Gm \Vdash \appInfAlg{A}{e} \sto \Om \Vdash \appInfAlg{A'}{e}[a][\jg']$
            and $\|\Om\| \vdash \appInf{A'}{e}{C}$.
\end{enumerate}
\end{corollary}

@@

The following corollary is derived immediately from Theorem~\ref{thm:completeness}.
\begin{corollary}[Completeness, single judgment form]
Given \emph{wf }$\Gm$ containing no judgment chains:
\begin{enumerate}
    \item If $\Om \vdash A' \le B'$ and $\Gm \Vdash A \le B \sto \Om \Vdash A' \le B'$
        \\then $\Gm \Vdash A \le B \redto \nil$;
    \item If $\Om \vdash e \Lto A'$ and $\Gm \Vdash e \Lto A \sto \Om \Vdash e \Lto A'$
        \\then $\Gm \Vdash e \Lto A \redto \nil$;
    \item If $\Om \vdash e \To A$ and $\Gm \Vdash e \To_a 1 \le 1 \sto \Om \Vdash e \To_a 1 \le 1$
        \\then $\Gm \Vdash e \To_a 1 \le 1 \redto \nil$;
    \item If $\Om \vdash \appInf{A'}{e}{C}$ and
        $\Gm \Vdash \appInfAlg{A}{e}[a][1 \to 1] \sto \Om \Vdash \appInfAlg{A'}{e}[a][1 \to 1]$
        \\then $\Gm \Vdash \appInfAlg{A}{e}[a][1 \le 1] \redto \nil$.
\end{enumerate}
\end{corollary}

@@

\begin{figure}[ht]
\centering\begin{tabular}{|c|c|c|l|}\hline
    File(s) & SLOC & \#Theorems & Description\\\hline
    olist.thm, nat.thm  &  309 & 56  & Basic data structures\\\hline
    typing.thm          &  238 & 5   & Declarative \& algorithmic system, debug examples\\\hline
    decl.thm            &  226 & 33  & Basic declarative properties\\\hline
    order.thm           &  235 & 27  & The $|\cdot|_\forall$ measure; decl. subtyping strenthening\\\hline
    alg.thm             &  655 & 72  & Basic algorithmic properties\\\hline
    trans.thm           &  628 & 53  & \makecell[l]{Transfer and its basic properties \\
                            Lemmas~\ref{lem:insert}, \ref{lem:extract}}\\\hline
    declTyping.thm      &  944 & 72  & \makecell[l]{Non-overlapping declarative system \\
                            Lemmas~\ref{lem:inv_allR}, \ref{lem:inv_chkAll}, \ref{lem:inv_chkLam}}\\\hline
    soundness.thm       & 1086 & 70  & Soundness theorem; aux. lemmas on transfer\\\hline
    depth.thm           &  206 & 14  & The $|\cdot|_\to$ measure; Lemma~\ref{lem:prune_inst}\\\hline
    dcl.thm             &  383 & 12  & Non-overlapping declarative worklist \\\hline
    completeness.thm    & 1075 & 61  &
                            Completeness theorem; aux. lemmas and relations\\\hline
    inst\_decidable.thm &  830 & 42  & Other measures; Lemma~\ref{lemma:inst:decidable}\\\hline
    decidability.thm    &  885 & 50  & Decidability theorem\\\hline
    Total               & 7700 & 567 & (51 definitions in total)\\\hline
\end{tabular}
\caption{Statistics for the proof scripts}
\label{fig:proof_statistics}
\end{figure}
